---
title: "SDS 4030 - Statistics for Data Science II: Project"
author: "Noah Calabrese"
subtitle: Main Markdown File
output:
  html_document:
    df_print: paged
  pdf_document: default
fontsize: 11pt
header-includes:
- \usepackage{hyperref}
- \hypersetup{pdfpagemode={UseNone}, pdfstartview=}
- \usepackage{amsthm}
- \theoremstyle{definition}
- \newtheorem{problem}{Problem}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4, fig.width = 5)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 68), tidy = TRUE)
knitr::opts_chunk$set(message = FALSE)
set.seed(0)
library(tidyverse)
#install.packages("tinytex")
#tinytex::install_tinytex()
library(tinytex)
#install.packages("formatR")
library(formatR)
```

Preliminary Reading Code In and library loading
```{r}
gss = read.table("gss.txt", header = TRUE)
#summary(gss)
#gss
```

Quick Cleaning of the Code. The Zodiac column has no relevance, it is not quantitatve and has no correlation to anything we need to regress so it is best if we just entirely remove it. It is an essential part of cleaning code. 
```{r}
#Removing Zodiac since it is not relevant in anyway to the analysis 
gss <- gss[-c(12)]
```

-------------------------------------------------------------------------------------------------------------------------------------------------------------------

Step 1: 

  i) Fit a multiple linear model with all available quantitative covariates. 
  
  ii) Perform exact and asymptotic tests for the set of linear constraints which has been specified for your team. 
  
  iii) Do all of your conclusions agree with each other and what are the relationships between the different tests?
  
-------------------------------------------------------------------------------------------------------------------------------------------------------------------

i) Fitting a multiple linear model with all available quantitative covariates. This is the model that we will use for the next few parts of this project. The reduced model is a little more tricky where we have to create a model in which it follows the constraints given by: $H_0 : 10 ∗ \beta_{M Educ} = 20 ∗ \beta_{F Educ} = 2$. So, we have to remove MEduc and FEduc since they are fixed parts of the response in the reduced model, so we remove them as predictors since they are specially accounted for.
```{r}
full_model <- lm(Income ~ Age + Educ + Hours + TVHours + MEduc + FEduc + Sibs, 
                   data = gss)

#Reduced model adhereing to the hypotheses constraints
gss$Income_adj <- gss$Income - (0.2 * gss$MEduc) - (0.1 * gss$FEduc)
reduced_model <- lm(Income_adj ~ Age + Educ + Hours + TVHours + Sibs, data = gss)
```


ii) We have to test the hypothesis $H_0 : 10 ∗ \beta_{M Educ} = 20 ∗ \beta_{F Educ} = 2$ against exact and asymptotic tests. The best exact test is, of course, the exact F-test assuming we are have normalcy conditions (which we will check later, but step 1 ensures that we do it regardless of the normalcy checks). To test for the asymptotic tests it is logical to use the Wald test since we have  a large sample (n = 815). It is also one of the easier asmyptotic tests to check, so we will use it due to its efficiency, accuracy, and sample size.
```{r}
#F-Exact Test
beta <- coef(full_model)                 
X <- model.matrix(full_model)            
n <- nrow(gss)                    
p <- length(beta)                
S2 <- summary(full_model)$sigma^2
R <- rbind(c(0, 0, 0, 0, 0, 1, 0, 0), 
           c(0, 0, 0, 0, 0, 0, 1, 0))
#the 1's in R represent the MEduc and FEduc since those are within the null hyp

#adjusting given the constraints from the null hypothesis
#2/10 = 0.2 | 1/10 = 0.1
r <- c(0.2, 0.1) 
k <- length(r)

FT = crossprod(R %*% beta - r, solve(R %*% solve(crossprod(X), t(R)), 
                                     R %*% beta - r))[,] / (k * S2)
print(FT)
anova(reduced_model, full_model)

#Wald Asymptotic Test
WT = k * FT
print(WT)
pvalue = pchisq(WT, k, lower.tail = FALSE)
anova(reduced_model, full_model, test = "Chisq")

```

iii) The conclusions agree with each other, both the F-Exact test and the Wald test provide very small test statistics (0.06 and 0.13 respectively), which means we do not have substantial evidence to reject the null hypothesis. There is no evidence suggesting that we should reject, even the p-values of each covariate in the anova test are at least above 0.08. To be specific, there is not a single statistically significant variable shown within these models. Also, the Wald Test is consistent with it's calculation of k X FT which is expected since n = 815 which is large enough to be accurate for an asymptotic test. That is further proof that the conclusions agree with each other.
-------------------------------------------------------------------------------------------------------------------------------------------------------------------

Step 2: 

  i) Is the variance of the random error terms non-constant as a function of any of the quantitative or qualitative covariates? 
  
  ii) Perform suitable heteroscedasticity tests and state your conclusions.
  
-------------------------------------------------------------------------------------------------------------------------------------------------------------------

i) After plotting a residuals vs fitted plot of the full model (shown in part 1), I think it is clear that this graph has non-constant variance which means that it likely suffers from heteroscedasticity. It also has some extreme outliers and leverage points (also called influential outliers). The points are basically pulling the regression line and likely causing that massive variance, or at least factoring into the magnitude of it.
```{r}
#Visualization of residuals against the fitted values 
plot(fitted(full_model), residuals(full_model), 
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red")

```

ii) The Breush-Pagan test for heteroscedasticity actually gives us a different answer than anticipated. From the BP test, we actually get the conclusion that we can not reject the null hypothesis that the full_model is homooscedastic since the p-value is larger than the alpha threshold of 0.05 (the p-value is at 0.23). This means we do not have sufficient evidence to reject the null hypothesis, so we cannot call the data heteroscedastic. This could be done because of the leverage outliers that is overriding the test, causing the test see random noise rather than a linear increase in variance, which leads to a high p-value. That is a possible explanation for this. 
```{r}
#Breusch-Pagan Test for heteroscedasticity
Y = gss$Income
residuals = full_model$residuals


Yaux = residuals^2
aux = lm(Yaux ~ Age + Educ + Hours + TVHours + MEduc + FEduc + Sibs, 
                   data = gss)
BP = n * summary(aux)$r.squared
print(BP)
pvalue = pchisq(BP, p, lower.tail = FALSE)
print(pvalue)

library(lmtest)
bptest(full_model)
```

-------------------------------------------------------------------------------------------------------------------------------------------------------------------

Step 3: 

  i) Calculate a suitable heteroscedasticity-consistent estimate for the covariance matrix of the regression coefficients. How does it compare against the
     corresponding ordinary least squares estimate? 
     
  ii) Calculate heteroscedasticity-consistent t-test statistics and 95 percent CIs for the regression coefficients. How do they compare against the results           yielded by the ordinary least squares method?

-------------------------------------------------------------------------------------------------------------------------------------------------------------------

i)
```{r}
#The third heteroscedasticity-consistent estimate is most robust


#Comparing Robust Standard Errors against the OLS Standard Errors


```

ii)
```{r}
#T-test for the heteroscedasticity-consistent regression coefficient


#95% CI for the heteroscedasticity-consistent regression coefficient


#Overall Comparison


```

