---
title: "SDS 4030 - Statistics for Data Science II: Final Project"
author: "Noah Calabrese, Natalia Jamula, Tim Mellman"
subtitle: Main Markdown File
output:
  pdf_document: default
  html_document:
    df_print: paged
fontsize: 11pt
header-includes:
- \usepackage{hyperref}
- \hypersetup{pdfpagemode={UseNone}, pdfstartview=}
- \usepackage{amsthm}
- \theoremstyle{definition}
- \newtheorem{problem}{Problem}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.height = 4, fig.width = 5)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 68), tidy = TRUE)
knitr::opts_chunk$set(message = FALSE)
set.seed(0)
library(tidyverse)
#install.packages("tinytex")
#tinytex::install_tinytex()
library(tinytex)
#install.packages("formatR")
library(formatR)
```

Preliminary library loading, reading code in, and exploratory data analysis

```{r}
gss = read.table("gss.txt", header = TRUE)
summary(gss)
```
We begin our data science workflow by familiarizing ourselves with the GSS dataset. The dataset contains 815 rows corresponding to 815 individuals' survery responses. The dataset contains 13 variables. There are 8 discrete numerical variables: Age (Responder's Age), Educ (Total years of education), Hours (Weekly work hours), TVHours (Hours spent watching TV daily), MEduc (Mother's years of education), FEduc (Father's years of education), Sibs (Number of siblings), and Income (Yearly income). Thus there are 5 categorical variables: Sex (Responder's sex), Happy (Happiness level), Marital (Marital status), Work (Work status), and Zodiac (Zodiac sign). Uniquely they take on the following values: Sex (Female, Male), Happy (Pretty Happy, Very Happy, Not Too Happy), Marital ("Divorced, Widowed, Married, Never Married, Separated), Work (Fulltime, Housekeeping, Laid Off, School, Retired, Parttime, Not Working, Other), and Zodiac (Virgo, Libra, Taurus, Pisces, Cancer, Aquarius, Leo, Capricorn, Aries, Gemini, Sagittarius, Scorpio). 

Now that we understand our data we isolate our goal being to build a linear model for the response variable yearly income by treating all other seven quanitative variables as covariates and ignoring all five qualitative variables. 

-------------------------------------------------------------------------------------------------------------------------------------------------------------------

Step 1: 

  i) Fit a multiple linear model with all available quantitative covariates. 
  
  ii) Perform exact and asymptotic tests for the set of linear constraints which has been specified for your team. 
  
  iii) Do all of your conclusions agree with each other and what are the relationships between the different tests?
  
-------------------------------------------------------------------------------------------------------------------------------------------------------------------

i) Fitting a multiple linear model with all available quantitative covariates. This is the model that we will use for the next few parts of this project

```{r}
library(faraway)
full_model <- lm(Income ~ Age + Educ + Hours + TVHours + MEduc + FEduc + Sibs, gss)
summary(full_model)
```

For the sake of later model comparison we will find the least squares estimate of the coefficient vector, an unbiased estimate of the residual variance and the MLE of the residual variance in the full model. 

```{r}
Y = gss$Income
X = model.matrix(full_model)
n = dim(X)[1]
p = dim(X)[2]
beta = solve(crossprod(X), crossprod(X, Y))[, ]
print(beta)
```

```{r}
residuals = Y - X %*% beta
sigma = sqrt(mean(residuals^2))
print(sigma^2)
```

```{r}
S = sqrt(sum(residuals^2)/(n - p))
print(S^2)
```

Our full model is $Y_i = \beta_0 +\beta_{Age}{\text{Age}_i} +\beta_{Educ}{\text{Educ}_i}+ +\beta_{Hours}{\text{Hours}_i} + \beta_{TVHours}{\text{TVHours}_i} + \beta_{MEduc}{\text{MEduc}_i} + \beta_{FEduc}{\text{FEduc}_i} + \beta_{Sibs}{\text{Sibs}_i} + \epsilon_i$. An unbiased estimate of the residual variance is equal to 5.42076e+12 and the MLE of the residual variance is equal to 5.474497e+12, under the full model. 

The reduced model is a little more tricky where we have to create a model in which it follows the constraints given by: $H_0 : 10 * \beta_{MEduc} = 20 * \beta_{FEduc} = 2$. 

We can rewrite the constraint as $10 * \beta_{MEduc} = 2$ , $20 * \beta_{FEduc} = 2$. 
Or, $\beta_{MEduc} = 0.2$, $\beta_{FEduc} = 0.1$

We can plug the null values into the full model, replacing coefficients with the constants so the full model becomes
 $Y_i = \beta_0 +\beta_{Age}{\text{Age}_i} +\beta_{Educ}{\text{Educ}_i} +\beta_{Hours}{\text{Hours}_i} + \beta_{TVHours}{\text{TVHours}_i} + 0.2*{\text{MEduc}_i} + 0.1*{\text{FEduc}_i} + \beta_{Sibs}{\text{Sibs}_i} + \epsilon_i$

We have to set MEduc and FEduc, the fixed parts of the response in the reduced model, as the offset. 
```{r}
reduced_model <- lm(Income ~ Age + Educ + Hours + TVHours + Sibs, gss, offset = 0.2*MEduc + 0.1*FEduc)
summary(reduced_model)
```
```{r}
R <- rbind(c(0, 0, 0, 0, 0, 1, 0, 0), 
           c(0, 0, 0, 0, 0, 0, 1, 0)) ##the 1's in R represent the MEduc and FEduc since those are within the null hypothesis
r <- c(0.2, 0.1) 
k <- length(r)
beta0 = beta - solve(crossprod(X), crossprod(R, solve(R %*% solve(crossprod(X),
t(R)), R %*% beta - r)))[, ]
print(beta0)
```

```{r}
residuals0 = Y - X %*% beta0
sigma0 = sqrt(mean(residuals0^2))
print(sigma0^2)
```

```{r}
S0 = sqrt(sum(residuals0^2)/(n - p + k))
print(S0^2)
```

An unbiased estimate of the residual variance is equal to 5.421657e+12 and the MLE of the residual variance is equal to 5.461868e+12, under the reduced model. 

ii) We have to test the hypothesis $H_0 : 10 * \beta_{MEduc} = 20 * \beta_{FEduc} = 2$ against exact and asymptotic tests. The best exact test is, of course, the exact F-test assuming we are have normalcy conditions (which we will check later, but step 1 ensures that we do it regardless of the normalcy checks). To test for the asymptotic tests it is logical to use the Wald test since we have a large sample (n = 815). It is also one of the easier asymptotic tests to check, so we will use it due to its efficiency, accuracy, and sample size.


### Exact F- Test 

```{r}
FT = crossprod(R %*% beta - r, solve(R %*% solve(crossprod(X), t(R)), 
                                     R %*% beta - r))[,] / (k * S^2)
print(FT)
```

```{r}
pvalue = pf(FT, k, n - p, lower.tail = FALSE)
print(pvalue)
```

```{r}
anova(reduced_model, full_model)
```
The observed value of the F test statistic is equal to 0.06680263 and the corresponding p-value is equal to 0.935385, so we fail to reject the null hypothesis at significance level $\alpha$ = 5%.

### Wald Test

```{r}
WT = crossprod(R %*% beta - r, solve(R %*% solve(crossprod(X), t(R)), R %*% beta - r))[, ]/S^2
print(WT)
```

```{r}
pvalue = pchisq(WT, k, lower.tail = FALSE)
print(pvalue)
```

```{r}
anova(reduced_model, full_model, test = "Chisq")
```
The observed value of the Wald test statistic is equal to 0.1336053 and the corresponding p-value is equal to 0.9353798, so we fail to reject the null hypothesis at significance level $\alpha$ = 5%

### Score Test

```{r}
SF = crossprod(X, residuals0)/sigma0^2
FI = crossprod(X)/sigma0^2
ST = crossprod(SF, solve(FI, SF))[, ]
print(ST)
```

```{r}
P = X %*% solve(crossprod(X), t(X))
ST = sum((P %*% residuals0)^2)/sigma0^2
print(ST)
```

```{r}
pvalue = pchisq(ST, k, lower.tail = FALSE)
print(pvalue)
```

The observed value of the Score test statistic is equal to 0.1349074 and the corresponding p-value is
equal to 0.934771, so we fail to reject the null hypothesis at significance level $\alpha$ = 5%. 

### Likelihood Ratio Test

```{r}
LR = -2 * (logLik(reduced_model)[1] - logLik(full_model)[1])
print(LR)
```

```{r}
LR = n * log(sigma0^2/sigma^2)
print(LR)
```

```{r}
pchisq(LR, k, lower.tail = FALSE)
```

```{r}
library(lmtest)
lrtest(reduced_model, full_model)
```
The observed value of the Likelihood Ratio test statistic is equal to 0.1349186 and the corresponding
p-value is equal to 0.9347658, so we fail to reject the null hypothesis at significance level $\alpha$ = 5%. 


iii) Yes, all of my conclusion agree with each other as in the exact F test, Wald test, score test, and likelihood ratio test we have failed to reject the null, and they all yield have similar p-values. Both the F-Exact test and asymptotic tests provide very small test statistics which means we do not have substantial evidence to reject the null hypothesis. 


```{r}
WT = k * FT
print(WT)
```

```{r}
ST = WT * S^2/sigma0^2
print(ST)
```

```{r}
LR = n * log(n/(n - p) * WT/ST)
print(LR)
```

```{r}
LR = n * log(1 + WT/(n - p))
print(LR)
```

```{r}
LR = -n * log(1 - ST/n)
print(LR)
```

In the F test, the exact F-statistic is 0.06680263, and the p-value is 0.935385. In the Wald test the Wald statistic is 0.1336053 and the p-value is 0.9353798. According to Slutsky’s theorem, we infer that $W_n$ = kF, the Wald statistic equals kF asymptotically. The Wald statistic is k times the 0.06680263 where k=2. While the F test relies on a finite sample, the Wald statistic follows a chi-square distribution and asymptotic approximations. 

In the score test, the score test statistic is 0.1349074 and the p-value is 0.934771 We know that $\hat{\sigma}_0^2 ST = S^2 W_n$. The Wald and Score test statistics only differ in terms of the residual variance estimation. The Wald test statistic utilizes an unbiased estimator of the residual variance under the full model while the Score test statistic utilizes the MLE of the residual variance under the reduced model. In other words the Wald test uses the full model variance and the Score test uses reduced model variance. Thus we can can see how the two test statistics differ slightly. 

In the likelihood ratio test, likelihood test statistic is 0.1349186 and the p-value is 0.9347658. Below we see how the Likelihood test statistic relates to the Wald test and Score test through transformations, implying they provide the same inference for very large sample sizes. 

So yes, we observe that the results of all the four test statistics concur. The results of the asymptotic Wald, Likelihood Ratio and Score tests are perfectly in line with the results of the exact F test, which is expected since n = 815 which is large enough to be accurate for an asymptotic test. All three asymptotic tests, Wald, Score & Likelihood Ratio Tests, are all asymptotically equivalent, converging to the same test statistics and giving us the same conclusions. They measure deviation from the null differently, where the Wald test measures the horizontal differences of the parameters and the null, the Score test measures the slope of the log likelihood  at the null, and the likelihood ratio test measures the vertical differences in log likelihood.  


-------------------------------------------------------------------------------------------------------------------------------------------------------------------

Step 2: 

  i) Is the variance of the random error terms non-constant as a function of any of the quantitative or qualitative covariates? 
  
  ii) Perform suitable heteroscedasticity tests and state your conclusions.
  
-------------------------------------------------------------------------------------------------------------------------------------------------------------------

i) We begin by plotting and examining a residuals vs fitted plot of the full model (shown in part 1). We can note that the residuals increase as the fitted values increases and also has some extreme outliers and leverage points (also called influential outliers). The points are basically pulling the regression line and likely causing that massive variance, or at least factoring into the magnitude of it.


```{r}
library(faraway)
full_model <- lm(Income ~ Age + Educ + Hours + TVHours + MEduc + FEduc + Sibs,
                 data = gss)
Y = gss$Income
X = model.matrix(full_model)
n = dim(X)[1]
p = dim(X)[2] - 1
residuals = full_model$residuals
plot(fitted(full_model), residuals, 
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red")
```


## Quantitative Covariates

### Responder's age 

```{r}
plot(rstandard(full_model) ~ Age, data = gss, ylab = "Standardized Residuals",
     xlab = "Age", pch = 16)
abline(h = 0, lty = 2, lwd = 2)
```

### Total years of education
```{r}
plot(rstandard(full_model) ~ Educ, data = gss, ylab = "Standardized Residuals",
     xlab = "Educ", pch = 16)
abline(h = 0, lty = 2, lwd = 2)
```

### Weekly work hours
```{r}
plot(rstandard(full_model) ~ Hours, data = gss, ylab = "Standardized Residuals",
     xlab = "Hours", pch = 16)
abline(h = 0, lty = 2, lwd = 2)
```

### Hours spent watching TV daily 
```{r}
plot(rstandard(full_model) ~ TVHours, data = gss, ylab = "Standardized Residuals",
     xlab = "TVHours", pch = 16)
abline(h = 0, lty = 2, lwd = 2)
```


### Mother's years of education

```{r}
plot(rstandard(full_model) ~ MEduc, data = gss, ylab = "Standardized Residuals",
     xlab = "MEduc", pch = 16)
abline(h = 0, lty = 2, lwd = 2)
```

### Father's years of education
```{r}
plot(rstandard(full_model) ~ FEduc, data = gss, ylab = "Standardized Residuals",
     xlab = "FEduc", pch = 16)
abline(h = 0, lty = 2, lwd = 2)
```

### Number of siblings
```{r}
plot(rstandard(full_model) ~ Sibs, data = gss, ylab = "Standardized Residuals",
     xlab = "Sibs", pch = 16)
abline(h = 0, lty = 2, lwd = 2)
```
After inspecting the standardized residuals plotted against each quantitative covariate, I noticed the presence of extreme outliers but no systematic pattern of increasing or decreasing spread as a function of a quantitative covariate that would indicate the form of heteroscedasticity that appears when the variance of the random error terms is a function of some of the available covariates.


## Qualitative Covariates

### Responder's sex

```{r}
boxplot(rstandard(full_model) ~ Sex, data = gss, ylab = "Standardized Residuals",
        xlab = "Sex")
abline(h = 0, lty = 2, lwd = 2)
```

### Happiness level
```{r}
boxplot(rstandard(full_model) ~ Happy, data = gss, ylab = "Standardized Residuals",
        xlab = "Happy")
abline(h = 0, lty = 2, lwd = 2)
```

### Marital status
```{r}
boxplot(rstandard(full_model) ~ Marital, data = gss, ylab = "Standardized Residuals",
        xlab = "Marital")
abline(h = 0, lty = 2, lwd = 2)
```

### Work status
```{r}
boxplot(rstandard(full_model) ~ Work, data = gss, ylab = "Standardized Residuals",
        xlab = "Work")
abline(h = 0, lty = 2, lwd = 2)
```

### Zodiac sign
```{r}
boxplot(rstandard(full_model) ~ Zodiac, data = gss, ylab = "Standardized Residuals",
        xlab = "Zodiac")
abline(h = 0, lty = 2, lwd = 2)
```


After inspecting the standardized residuals plotted against each qualititative covariate, I again noticed the presence of extreme outliers but no pattern of difference in spread across the levels of the qualitative covariate  that would indicate the form of heteroscedasticity that appears when the variance of the random error terms differs across the different levels of some qualitative covariate. 

So no, we do not have strong evidence to conclude that the variance of the random error terms is non-constant as a function of
any of the quantitative or qualitative covariates.

ii) We use the Breush-Pagan test for heteroscedasticity to test whether the variance of the error terms is a lineae function of the residuals. The BP test statistic is 8.955868 and p-value is 0.2558401. Thus, we can not reject the null hypothesis that the full_model is homoscedastic since the p-value is larger than the alpha threshold of 0.05. This means we do not have sufficient evidence to reject the null hypothesis of homoskedasticity at $\alpha$ = 5% and we do not detect a relationship of linear dependence of variance on the covariates. 


```{r}
#Breusch-Pagan Test for heteroscedasticity
Y = gss$Income
residuals = full_model$residuals
Yaux = residuals^2
aux = lm(Yaux ~ Age + Educ + Hours + TVHours + MEduc + FEduc + Sibs, 
                   data = gss)
BP = n * summary(aux)$r.squared
print(BP)
```

```{r}
pvalue = pchisq(BP, p, lower.tail = FALSE)
print(pvalue)
```

```{r}
library(lmtest)
bptest(full_model)
```

-------------------------------------------------------------------------------------------------------------------------------------------------------------------

Step 3: 

  i) Calculate a suitable heteroscedasticity-consistent estimate for the covariance matrix of the regression coefficients. How does it compare against the corresponding ordinary least squares estimate? 
     
  ii) Calculate heteroscedasticity-consistent t-test statistics and 95 percent CIs for the regression coefficients. How do they compare against the results yielded by the ordinary least squares method?

-------------------------------------------------------------------------------------------------------------------------------------------------------------------


```{r}
X = model.matrix(full_model)
n = dim(X)[1]
p = dim(X)[2]
beta = full_model$coefficients
residuals = full_model$residuals
vcovOLS = summary(full_model)$sigma^2*summary(full_model)$cov.unscaled
print(vcovOLS)
```
The covariance matrix of the least squares estimator is $\begin{pmatrix} 474220174987 & -1907433705.2 & -12064670132 & -1347058849.0 & -8426312775 & -5079147869.7 & -4109118474 & -9103007619 \\  -1907433705 & 38829681.0 & -31202461 & -432037.2 & -15594169 & 29086917.4 & 37000089 & -7167222 \\  -12064670132 & -31202461.2 & 1269225187 & -20945952.6 & 200001770 & -236010377.3 & -154559694 & 167953096 \\  -1347058849 & -432037.2    & -20945953 & 35081614.4 & 15414790 & 138585.3 & 6151805 & 14433306 \\ -8426312775 & -15594168.6 & 200001770 & 15414789.6 & 1805687552 & 28954807.1 & 53859762 & -56165946 \\ -5079147870 & 29086917.4 & -236010377 & 138585.3 & 28954807 & 1069222108.8 & -493576358 & 79738958 \\ -4109118474 & 37000088.9 & -154559694 & 6151805.3  & 53859762 & -493576358.5 & 825093313 & 159284579 \\ -9103007619 & -7167222.2 & 167953096 & 14433306.1 & -56165946 & 79738957.8 & 159284579 & 1121549496 \end{pmatrix}$. 


```{r}
confint(full_model)
```
The 95% observed CI for the regression coefficient $\hat\beta_0$, the intercept, is [-1139787.136, 1563672.818]. The 95% observed CI for the regression coefficient$\hat\beta_{Age}$, is [-9282.145, 15180.975]. The 95% observed CI for the regression coefficient$\hat\beta_{Educ}$, is [-49482.462, 90379.388]. The 95% observed CI for the regression coefficient$\hat\beta_{Hours}$, is [-20949.732, 2302.771]. The 95% observed CI for the regression coefficient$\hat\beta_{TVHours}$, is [-148773.710, 18047.486]. The 95% observed CI for the regression coefficient$\hat\beta_{MEduc}$, is [-67900.262, 60469.842].The 95% observed CI for the regression coefficient$\hat\beta_{FEduc}$, is [-46177.946, 66588.907]. The 95% observed CI for the regression coefficient$\hat\beta_{Sibs}$, is [-7087.921, 124385.860].

```{r}
summary(full_model)
```
The observed t test statistic of statistical significance for the intercept regression coeﬃcient
is 0.308 with p-value 0.7583. The observed t test statistic of statistical significance for the Age regression coefficient is 0.473 with p-value 0.6361. The observed t test statistic of statistical significance for the Educ regression coefficient is 0.574 with p-value 0.5661. The observed t test statistic of statistical significance for the Hours regression coefficient is -1.574 with p-value 0.1159. The observed t test statistic of statistical significance for the TVHours regression coefficient is -1.538 with p-value 0.1244. The observed t test statistic of statistical significance for the MEduc regression coefficient is -0.114 with p-value 0.9096. The observed t test statistic of statistical significance for the FEduc regression coefficient is 0.355 with p-value 0.7225. The observed t test statistic of statistical significance for the Sibs regression coefficient is 1.751 with p-value 0.0803. Thus, none of the regression coefficients are statistically significant at the 0.05 level. 


i) Calculate a suitable heteroscedasticity-consistent estimate for the covariance matrix of the regression coefficients. How does it compare against the corresponding ordinary least squares estimate?

```{r}
Sigma = diag(residuals^2 / (1 - hatvalues(full_model))^2)
vcovHC3 = summary(full_model)$cov.unscaled %*% crossprod(X, Sigma) %*% X %*% summary(full_model)$cov.unscaled
print(vcovHC3)
```
The heteroscedasticity-consistent estimate for the covariance matrix of the regression coeﬃcients is $\begin{pmatrix} 647905975588 & -2119149267 & -19837922241 & -1357649949 & -9486014566 & 3595276136 & -17028391209 & -11637769372 \\ -2119149267 & 48814436 & -24590704 & -17494011 & -50260968 & -34217774    & 162045871 & -37033183 \\ -19837922241 & -24590704 &  1162078662 & 31381166 & 143728482 & -53411133 & 94715693 & 888805838 \\ -1357649949  & -17494011 & 31381166  & 27278306 & 103955035& 9117008 & -7441799 & -9925909 \\ -9486014566 & -50260968 & 143728482 & 103955035 & 1240245743&  30651678 & 103186441 & -486253897 \\ 3595276136 & -34217774 & -53411133 & 9117008 & 30651678 & 245096719  & -386380314 & -132459705 \\ -17028391209 & 162045871   & 94715693 & -7441799 & 103186441 & -386380314 & 1151032066 & 101992572 \\ -11637769372 & -37033183 & 888805838 & -9925909 & -486253897 & -132459705 & 101992572 &  1425697587 \end{pmatrix}$. 

The variance of the intercept increases from 474220174987 to 647905975588. The variance of the Age coefficient decreases from -1907433705.2 to -2119149267. The variance of the Educ coefficient decreases from -12064670132 to -19837922241.  The variance of the Hours coefficient decreases slightly from -1347058849.0 to -1357649949. The variance of the TVHours coefficient decreases from -8426312775 to -9486014566. The variance of the MEduc coefficient increases from -5079147869.7 to 3595276136. The variance of the FEduc coefficient increases from -4109118474 to -17028391209. The variance of the Sibs coefficient decreases from -9103007619 to -11637769372. 

So, the heteroscedasticity-consistent estimator overall produces broadly similar estimated variances and standard errors than the ordinary least squares estimator. The heteroscedasticity-consistent estimate for the covariance matrix of the least squares estimator is similar compared to the one obtained through the least squares method due to the lack of evidence for heteroscedasticity in the random error terms.

ii) Calculate heteroscedasticity-consistent t-test statistics and 95 percent CIs for the regression coefficients. How do they compare against the results yielded by the ordinary least squares method?

## Heteroscedasticity-consistent t-test statistics

```{r}
library(lmtest)
coeftest(full_model, vcov. = vcovHC3)
```
For the intercept coefficient, the heteroscedastisticity-consistent t test statistic is 0.2633, and corresponding p-value is 0.79238. Compared to the t-statistic and p-value yielded by the ordinary least squares method, 0.308 and 0.7583, respectively, the heteroscedasticity-consistent t-test statistic is slightly smaller with a slightly larger p-value. 

For the Age coefficient, the heteroscedastisticity-consistent t test statistic is 0.4221, and corresponding p-value is 0.67303. Compared to the t-statistic and p-value yielded by the ordinary least squares method, 0.473 with p-value 0.6361, respectively, the heteroscedasticity-consistent t-test statistic is slightly smaller with a slightly larger p-value. 

For the Educ coefficient, the heteroscedastisticity-consistent t test statistic is 0.5999, and corresponding p-value is 0.54877 Compared to the t-statistic and p-valye yielded by the ordinary least squares method, 0.574 with p-value 0.5661, respectively, the heteroscedasticity-consistent t-test statistic is slightly larger with a slightly smaller p-value. 

For the Hours coefficient, the heteroscedastisticity-consistent t test statistic is -1.7851, and corresponding p-value is 0.07462 Compared to the t-statistic and p-value yielded by the ordinary least squares method, -1.574 with p-value 0.1159, respectively, the heteroscedasticity-consistent t-test statistic is slightly more negative with a slightly smaller p-value. 

For the TVHours coefficient, the heteroscedastisticity-consistent t test statistic is -1.8560, and corresponding p-value is 0.06382 Compared to the t-statistic and p-value yielded by the ordinary least squares method, -1.538 with p-value 0.1244, respectively, the heteroscedasticity-consistent t-test statistic is slightly more negative with a slightly smaller p-value. 

For the MEduc coefficient, the heteroscedastisticity-consistent t test statistic is -0.2373, and corresponding p-value is 0.81248. Compared to the t-statistic and p-value yielded by the ordinary least squares method, -0.114 with p-value 0.9096, respectively, the heteroscedasticity-consistent t-test statistic is slightly more negative with a slightly smaller p-value. 

For the FEduc coefficient, the heteroscedastisticity-consistent t test statistic is 0.3008, and corresponding p-value is 0.76364 Compared to the t-statistic and p-value yielded by the ordinary least squares method, 0.355 with p-value 0.7225, respectively, the heteroscedasticity-consistent t-test statistic is slightly smaller with a slightly larger p-value.

For the Sibs coefficient, the heteroscedastisticity-consistent t test statistic is 1.5533, and corresponding p-value is 0.12075 Compared to the t-statistic and p-value yielded by the ordinary least squares method, 1.751 with p-value 0.0803, respectively, the heteroscedasticity-consistent t-test statistic is slightly smaller with a larger p-value.

Both methods do not produce statistically significant p-values for all coefficients.

## 95 percent CIs

```{r}
alpha = 0.05
CI = cbind(beta - qt(1 - alpha/2, n - p) * sqrt(diag(vcovHC3)),
           beta + qt(1 - alpha/2, n - p) * sqrt(diag(vcovHC3)))
colnames(CI) = colnames(confint(full_model))
print(CI)
```
The heteroscedastisticity-consistent 95% CIs for the intercept coefficient is [-1368052.93, 1791938.6124]. Compared to the results yielded by the ordinary least squares method [-1139787.136, 1563672.818], the heteroscedastisticity-consistency interval is much wider. 

The heteroscedastisticity-consistent 95% CIs for the Age coefficient is [-10764.90, 16663.7286]. Compared to the results yielded by the ordinary least squares method [-9282.145, 15180.975], the heteroscedastisticity-consistency interval is wider. 

The heteroscedastisticity-consistent 95% CIs for the Educ coefficient is [-46465.65, 87362.5712]. Compared to the results yielded by the ordinary least squares method [-49482.462, 90379.388], the heteroscedastisticity-consistency interval is much wider.

The heteroscedastisticity-consistent 95% CIs for the Hours coefficient is [-19575.48, 928.5205]. Compared to the results yielded by the ordinary least squares method [-20949.732, 2302.771], the heteroscedastisticity-consistency interval is actually not wider. 

The heteroscedastisticity-consistent 95% CIs for the TVHours coefficient is [-134491.08, 3764.8580]. Compared to the results yielded by the ordinary least squares method [-148773.710, 18047.486], the heteroscedastisticity-consistency interval is much wider.

The heteroscedastisticity-consistent 95% CIs for the MEduc coefficient is [-34445.64, 27015.2230]. Compared to the results yielded by the ordinary least squares method [-67900.262, 60469.842], the heteroscedastisticity-consistency interval is much more narrow. 

The heteroscedastisticity-consistent 95% CIs for the FEduc coefficient is [-56389.83, 76800.7903]. Compared to the results yielded by the ordinary least squares method [-46177.946, 66588.907], the heteroscedastisticity-consistency interval is more narrow. 

The heteroscedastisticity-consistent 95% CIs for the Sibs coefficient is [-15467.31, 132765.2539]. Compared to the results yielded by the ordinary least squares method [-7087.921, 124385.860], the heteroscedastisticity-consistency interval is much wider.

The heteroscedastisticity-consistent 95% CIs for all regression coeﬃcients besides MEduc and FEduc are wider than the 95% CIs with by the ordinary least squares method, but all methods contain 0 so the conclusions are the same at the 5% level and all regression coefficients are not statistically significant.  



